Detecting diversifying selection for a trait from within and between-species genotypes and phenotypes

### 2024-07-05 manuscript/bayescode_nodetraits.tex
Cleaned unecessary files. Updated README.

\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{1.2}
\linenumbers
\frenchspacing

@@

\maketitle

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To quantify selection acting on a trait, methods have been developed using either within or between-species variation.
    However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
    Conversely, current methods using between-species variation usually discard within-species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

\keywords{Quantitative genetics \and Trait evolution \and Selection \and Phylogenetics \and Population genetics }

% Research Article (7500 words)
\section*{Introduction}\label{sec:introduction}

Determining whether a trait is under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\citep{lynch_genetics_1998, walsh_evolution_2018}.

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait) and under stabilizing selection, while the loci affecting those traits are mostly pleiotropic (many traits associated with a given locus) with additive effects~\citep{simons_population_2018, sella_thinking_2019}.
Given this genetic architecture of traits, from two diverging populations, it is possible to distinguish which traits have evolved under natural selection in controlled experimental settings, by performing genetic cross between individuals~\citep{fraser_detecting_2020}.
Across several populations, by contrasting both trait differentiation (\Qst) and genetic differentiation (\Fst), so-called \QstFst\ methods have been used to determine the selective regime and to quantify the strength of selection acting on a trait~\citep{merila_comparison_2001, leinonen_comparative_2008}.
\Qst\ higher than \Fst\ is interpreted as a signature of diversifying selection due to adaptation to different optimum trait values in the different populations.
Contrarily, \Qst\ lower than \Fst\ is interpreted as a signature of stabilizing selection~\citep{lamy_qst_2012}.
Other frameworks explicitly model genetic drift as a random process generating both trait and genetic differences between individuals and populations.
This integrated framework can discriminate between selection and genetic drift as a cause of trait differentiation between populations of the same species~\citep{ovaskainen_new_2011}.
However, regardless of the strengths and weaknesses of each method~\citep{pujol_are_2008, edelaar_comparisons_2011, ovaskainen_new_2011}, tests of trait differentiation between populations are ultimately limited to recent local adaptation since they are based on the variation observed within a single species.

% Pair of species
To disentangle selection from neutral evolution, trait variation can also be observed at a larger time scale.
For example, starting from the same ancestral population, divergent lineages accumulate phenotypic changes that will reach fixation in the population.
These changes ultimately result in different mean trait values across lineages.
Theoretically, the variance in mean trait value (between lineages) does increase linearly with time of divergence, and also proportionally to the trait variance at the population scale~\citep{lande_genetic_1980, turelli_heritable_1984, felsenstein_phylogenies_1988a}.
Empirically, this effect can be observed for genes with larger within-species variation in gene expression level, which exhibits a faster accumulation of divergence in mean expression level~\citep{khaitovich_neutral_2004}.
As an analogy, in the context of protein-coding DNA sequences, leveraging within species variation and divergence to a sister species is the crux of the \citet{mcdonald_adaptative_1991} test.
In such a test, inflation of divergence to the sister species is compared to polymorphism within species, while neutral makers (usually synonymous sites) are used to determine the neutral expectation and thus are used for normalization.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\citep{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\citep{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in mean trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\citep{felsenstein_phylogenies_1985, felsenstein_phylogenies_1988a, hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\citep{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Deviations from the assumptions of the Brownian process are however well known.
When trait variation is constrained because of optimum mean trait values across or between species, the pattern of evolution can be modeled by the Ornstein-Uhlenbeck processes, which is often interpreted as a signature of stabilizing selection~\citep{hansen_stabilizing_1997, catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction without fixed optimum) is interpreted as a signature of directional selection at the phylogenetic scale~\citep{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\citep{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also changing as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\citep{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Second, a better fit of a Brownian could be due to a trait evolving with a rate too low compared to the timespan on which it is measured~\citep{grabowski_cautionary_2023}.
And third, even for a trait evolving under a neutral regime, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\citep{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro-evolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation together with changes in mean trait value to either describe measurement errors~\citep{lynch_methods_1991, hansen_interpreting_2012}, incorporate values for individuals~\citep{felsenstein_comparative_2008} or to scale the rate of change in mean trait value~\citep{kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
Across many species, within-species variation has also been used to infer diversifying selection by estimating the ratio of between to within species variation of many traits and test for deviation from the average ratio across traits~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Here, our goal was again to use both variances between and within species to determine the selective regime of a quantitative trait.
We build a novel framework that integrates trait variation at the phylogenetic and population scales together with estimates of nucleotide sequence variations at both scales.
It allowed us to define an expected ratio of normalized variance between and within species while setting the threshold of this ratio for neutral, stabilizing, and diversifying selection.
The ratio that we propose can be considered as a neutrality index for any quantitative trait ~\citep{lynch_rate_1990}, while articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between species).
From the field of population genetics, while \QstFst\ methods and their derivatives ultimately seek trait differentiation among different populations from the same species~\citep{pujol_are_2008, ovaskainen_new_2011}, our study can be seen as their macro-evolutionary analog to account for phylogenetic relationships between species.
From the field of phylogenetics, our study can be seen as an alternative to the EVE model~\citep{rohlfs_modeling_2014, rohlfs_phylogenetic_2015} for a single trait, where we set a threshold for neutral evolution by leveraging species nucleotide polymorphism and divergence.

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

While observing trait variations across individuals of several species, we ask if the variation within species compared to variation between species is compatible with neutral evolution or not.
In statistical terms, this can also be framed as: Is the variance of means equal to the mean of variances?
The difficulty in such a study is that individuals are not independent samples, but are from species that diverged at different times.
By reviewing theoretical expectations and leveraging nucleotide sequence variations, the goal of this section is thus to obtain normalized trait variation between and within species that are equal if the trait is neutral.
Here we denote these normalized trait variations as respectively $\RateWhithin$ for within species and as $\RateBetween$ for between species.

\subsubsection*{Within-species trait variations}

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
For a diploid individual, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \citet{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per locus per generation ($\MutationRatePheno$) and the genetic architecture of the trait as:
\begin{gather}
    \VarMutation = 2 \MutationRatePheno \Multiply \GenArchi \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\citep{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\citep{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is the average number of differences between pairs of sequences drawn at random, which is also equal to the sum of expected heterozygosities over all nucleotide sites~\citep{tajima_statistical_1989}.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \citet{tajima_statistical_1989}, $\pi$ is a function of the mutation rate ($\MutationRateNuc$, per nucleotide site per generation) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRateNuc \label{eq:genetic-diversity}.
\end{gather}

To remove the effect of $\Ne$, we define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
After simplification, $\RateWhithin$ is then solely a function of the underlying genetic architecture as:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi} \label{eq:def-rate-pop}, \\
    & = \frac{4 \Ne \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \Ne \Multiply \MutationRateNuc} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-pop}
\end{align}

If $\VarGenetic$ is not empirically accessible, it can be related to the observed phenotypic variance ($\VarPhenotype$), multiplied by narrow-sense heritability of the trait ($\Heritability$), as~\citep{hill_data_2008}:
\begin{align}
    \VarGenetic & = \Heritability \Multiply \VarPhenotype.\label{eq:var-gen-pheno}
\end{align}
Which leads to $\RateWhithin$ being a function of $\VarPhenotype$ and $\Heritability$ instead of $\VarGenetic$ as:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi } \text{ from definition eq.~\ref{eq:def-rate-pop} and~\ref{eq:var-gen-pheno}}, \label{eq:def-rate-pheno-pop} \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \text{ from eq.~\ref{eq:rate-pop}} \label{eq:rate-pheno-pop}
\end{align}

\subsubsection*{Between-species trait variations}\label{subsec:between-species-var}
For a given species $\Spi$, we denote by $\MeanTrait_{\Spi}$ the mean value of the trait across the individuals.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait_{\Spi}$ evolves as a Brownian process~\citep{felsenstein_phylogenies_1985, hansen_translating_1996}.
Given a phylogenetic tree, for a pair of species $\Spi$ and $\Spj$ from this tree, we denote as $\NbrGen$ the number of generations between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
Then, the covariance between $\MeanTrait_{\Spi}$ and $\MeanTrait_{\Spj}$ depends on $\NbrGen$ as given in \citet{hansen_translating_1996} by:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \NbrGen \\
    & = 4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}}.
\end{align}
Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$~\citep{kimura_probability_1962}.
We can derive the substitution rate ($\SubRate$, per nucleotide site per generation) as the number of newly arisen mutations ($2\Ne \Multiply \MutationRateNuc$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\citep{kimura_evolutionary_1968}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRateNuc \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRateNuc  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRateNuc. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution per generation within a genomic region equals the rate at which new mutations arise per generation for the same genomic region~\citep[review]{mccandlish_modeling_2014}.

Next, we denote $\NucDiv$ as the nucleotide divergence between the root of the tree and the most recent common ancestor of taxa $\Spi$ and $\Spj$.
In other words, $\NucDiv$ is the expected number of substitutions per nucleotide site during the $\NbrGen$ generations.
Assuming that no multiple substitutions occurred at the same site, $\NucDiv$ is the number of generations ($\NbrGen$) multiplied by the nucleotide substitution rate per generation ($\SubRate$):
\begin{align}
    \NucDiv & = \NbrGen \Multiply \SubRate \\
    & = \NbrGen \Multiply \MutationRateNuc \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
To remove the effect of the number of generations ($\NbrGen$) first, and to also equate to $\RateWhithin$ (eq.~\ref{eq:rate-pop}), we define $\RateBetween$ as the covariance in the mean trait value ($\VarPhy$) normalized by 4 times the nucleotide divergence of any neutral genomic region ($4\NucDiv$).
After simplification, $\RateBetween$ is also solely a function of the underlying genetic architecture as:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 \NucDiv} \label{eq:def-rate-phy},\\
    & = \frac{4 \NbrGen \Multiply \MutationRatePheno \Multiply \GenArchi}{4 \NbrGen \Multiply \MutationRateNuc}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc}. \label{eq:rate-phy}
\end{align}

In equation~\ref{eq:rate-phy}, we show that the covariance in mean trait value between a pair of species ($\VarPhy$) does increase linearly with shared nucleotide divergence ($\NucDiv$), if the trait and sequences are neutrally evolving and the genetic architecture of the trait has not changed.
Importantly, since the number of generations is the ratio of time divided by generation time (average time between two consecutive generations), removing the effect of the number of generations in equation~\ref{eq:rate-phy} also removes the effect of both time and generation time.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure1}
    \caption{
        Between species, the change along the phylogeny of the mean phenotypic trait allows the estimation of between-species trait variation, $\RateBetween$, which is normalized by nucleotide divergence.
        Within species, the genetic variance allows the estimation of within-species trait variation, $\RateWhithin$, which is  normalized by nucleotide diversity.
        $\NI$ is the ratio of $\RateBetween$ over $\RateWhithin$.
        Under neutral evolution, $\NI$ is expected to be equal to one.
        Under diversifying selection, the trait is heterogeneous between species, but homogeneous within species, leading to $\NI$ greater than one.
        Under stabilizing selection, the trait is homogeneous between species, leading to $\NI$ smaller than one.
        Importantly, the sequence from which nucleotide diversity and divergence are estimated should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study, they allow for discarding the confounding effect on mutation rate diversity, population size and divergence time.
    }
    \label{fig:methods}
\end{figure*}

\subsubsection*{Neutrality index}

The variability between either individuals or species can be obtained for both quantitative traits and genomic sequences.
At the population level, the variability of the trait between individuals can be combined with the nucleotide diversity of any neutrally evolving genomic region to obtain $\RateWhithin$.
At the phylogenetic level, the variability of the mean trait value between species can be combined with the nucleotide divergence of any neutrally evolving genomic region to obtain $\RateBetween$.
If the trait is neutrally evolving and the genetic architecture of the trait has not changed along the phylogenetic tree, we thus have:
\begin{align}
    \frac{\RateBetween}{\RateWhithin} &= \frac{\VarPhy}{4 \NucDiv} \Multiply \frac{\pi}{\Heritability \Multiply \VarPhenotype} \text{ by definition and eq.~\ref{eq:def-rate-pheno-pop} and~\ref{eq:def-rate-phy}},\\
    &= \frac{\MutationRatePheno \Multiply \GenArchi}{\MutationRateNuc} \Multiply \frac{\MutationRateNuc}{\MutationRatePheno \Multiply \GenArchi} \text{ from eq.~\ref{eq:rate-pheno-pop} and~\ref{eq:rate-phy}}, \\
    &= 1.
\end{align}
We define a neutrality index $\NI$ as:
\begin{align}
\NI \defEqual \frac{\RateBetween}{\RateWhithin},
\end{align}
 which will equal to $1$ for a trait evolving neutrally.
Both $\RateBetween$ and $\RateWhithin$ can be estimated using quantitative trait and genomic sequences within and between species, while neither the mutation rates ($\MutationRatePheno$ and $\MutationRateNuc$), nor the effective population size ($\Ne$), generation time or time of divergence ($\NbrGen$) need to be estimated.
Moreover, the nucleotide sequence from which $\pi$ and $\NucDiv$ are obtained should be neutrally evolving, but they are not necessarily linked to the quantitative trait under study.

\subsection*{Estimation}\label{subsec:estimate}

We hereby seek to obtain point estimates of $\RateBetween$, $\RateWhithin$ and ultimately $\NI$.
For each species with data available, $\RateWhithin$ as defined in eq.~\ref{eq:def-rate-pheno-pop} can be seen as a replicate sample.
Thus, $\RateWhithin$ can be obtained by averaging out across all the sampled species.
On the other hand, $\RateBetween$ such as as defined in eq.~\ref{eq:def-rate-phy} only refers to a pair of species, and thus must be generalized to account for different species divergence, as is done in the comparative framework~\citep{felsenstein_phylogenies_1985, omeara_testing_2006}.
Generally, $\RateBetween$ can thus be seen as an estimate of the rate of evolution of the quantitative trait along a phylogenetic tree, when the tree is measured in units of $4d$ ($d$ being the nucleotide divergence).
As such, any phylogenetic comparative methods that allow the estimation of phenotypic rates of evolution on a tree scaled by $4d$, instead of time as is usually the case, can be used to estimate $\RateWhithin$.
We provide a maximum likelihood estimate for $\NI$ as well as a Bayesian estimate to derive posterior probabilities that the null model of neutrality (i.e.~$\NI = 1$) is rejected.

\subsubsection*{Maximum likelihood estimate}
At the phylogenetic scale, for $\NbrTaxa$ taxa in the tree, $\DistanceMatrix$ ($\NbrTaxa \times \NbrTaxa$) is the symmetric distance matrix computed from the branch lengths and the topology of the phylogenetic tree.
The diagonal $\Distance_{\Spi,\Spi}$ represents the total nucleotide divergence from the root of the tree to each taxon ($\Spi$).
The off-diagonal elements ($\Distance_{\Spi,\Spj} = \NucDiv$) are the distances between the root and the most recent common ancestor of taxa $\Spi$ and $\Spj$, as in equation~\ref{eq:distance}.
The mean trait value at the root of the tree ($\RootTrait$) can be estimated from the $\NbrTaxa \times 1$ vector of mean trait values $\VecTrait$ at the tips of the tree using maximum likelihood~\citep{omeara_testing_2006}:
\begin{gather}
    \RootTrait = \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecOne \right)\inv \Multiply \left( \VecOne\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \VecTrait \right), \label{eq:estimated-root-trait}
\end{gather}
where $\VecOne$ is an $\NbrTaxa \times 1$ column vector of ones.

Finally, between-species variation $\RateBetween$ is estimated as~\citep{omeara_testing_2006}:
\begin{gather}
    \RateBetween = \frac{1}{4}\frac{\left( \VecTrait -  \RootTrait \Multiply \VecOne \right)\tr \MultiplyMatrix \DistanceMatrix\inv \MultiplyMatrix \left( \VecTrait -  \RootTrait \Multiply \VecOne  \right)}{\NbrTaxa - 1}. \label{eq:estimated-rate-phy}
\end{gather}

For a given species $\Spi$ with inter-individual data available, additive genetic variance of a trait ($\VarGeneticSpi$) is the product of heritability ($\Heritability_{i}$) and phenotypic variance ($V_{\Trait, i}$).
The ratio of $\VarGeneticSpi$ over nucleotide diversity of neutrally evolving sequences ($\pi_{\Spi}$) is a sample estimate of $\RateWhithin$.
Averaged across all species, we obtain the estimate $\RateWhithin$ as:
\begin{gather}
    \RateWhithin = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa}\frac{  \VarGeneticSpi}{ \pi_{i}} = \dfrac{1}{\NbrTaxa}\sum_{i=1}^{\NbrTaxa} \frac{  V_{\Trait, i} \Multiply \Heritability_{i}}{ \pi_{i}}. \label{eq:estimated-rate-pop}
\end{gather}

As depicted in fig.~\ref{fig:methods}, the neutrality index is estimated as:
\begin{gather}
    \NI = \frac{\RateBetween}{\RateWhithin}. \label{eq:estimated-NI}
\end{gather}

\subsubsection*{Multivariate Brownian process}

In the previous section, $\NI$ is estimated independently for each trait of interest.
Here we generalize to $\Ntrait$ traits co-varying along the phylogenetic tree, since simultaneously estimating all $\RateBetween$ allows improving their estimation~\citep{adams_multivariate_2018}.
More specifically, trait variation along the phylogenetic tree is modeled as a $\Ntrait$-dimensional Brownian process $\Brownian$ ($1 \times \Ntrait$) starting at the root and branching along the tree topology~\citep{huelsenbeck_detecting_2003, lartillot_phylogenetic_2011, lartillot_joint_2012, latrille_inferring_2021}.
The rate of change of the Brownian process is determined by the positive semi-definite and symmetric covariance matrix between traits $\CovarianceMatrix$ ($\Ntrait \times \Ntrait$).
The branch lengths of the tree used to model the Brownian process runs is measured in units of $4d$ ($d$ being the nucleotide divergence).
The off-diagonal elements of $\CovarianceMatrix$ are the covariance between traits, and the diagonal elements are the variance of each trait when measured in $4d$ units, and thus equate to $\RateBetween$ (see section~S\ref{subsec:multivariate-brownian-process}).
Of note, modeling trait evolution as a multi-dimensional process is reliable only if $\Ntrait \ll \NbrTaxa$, meaning that the number of species is largely superior to the number of traits~\citep{adams_multivariate_2018}.
Thus, relying on a $\Ntrait$-dimensional process should be reserved for a handful of allometric traits (e.g. brain mass and body mass).
If $\Ntrait$ is large, the traits are better tested independently each with a $1$-dimensional Brownian process, which is a specific case of the multi-dimensional process.

\subsubsection*{Bayesian estimate}
The Bayesian framework allows obtaining the posterior distribution of neutrality index ($\NI$) for traits of interest.
We used the \textit{BayesCode} software to model $\Ntrait$-dimensional Brownian processes along a phylogenetic tree~\citep{latrille_inferring_2021}.
With an inverse Wishart distribution as the {prior} on the covariance matrix, the {posterior} on $\CovarianceMatrix$, conditional on $\brownian$ is also an invert Wishart distribution (see section~S\ref{subsec:sampling-the-covariance-matrix}).
We used Metropolis-Hastings algorithm to sample $\Brownian$, while the posterior distribution of $\CovarianceMatrix$ is sampled using Gibbs sampling.
For each trait and each species, the prior on heritability ($\Heritability$) for each species is set as a uniform distribution with user-defined boundaries.
Heritability and phenotypic variance for each trait are combined with nucleotide diversity to compute $\RateWhithin$ for each species before being averaged across species (as in eq.~\ref{eq:estimated-rate-pop}).
From $\RateWhithin$ estimated independently for each trait and the diagonal elements of $\CovarianceMatrix$ (i.e. the $\RateBetween$ for each trait), the posterior distribution of $\NI$ (as in eq.~\ref{eq:estimated-NI}) is obtained for each trait.
The posterior distribution of $\NI$ thus allows testing for deviation from neutrality (Fig.~\ref{fig:methods}), for example, by computing $\proba [\NI > 1 ]$ to test for evidence of diversifying selection and $\proba [\NI < 1 ]$ to test for evidence of stabilizing selection.

\subsubsection*{Applicability to empirical data}

Our method assumes that the narrow-sense heritability ($\Heritability$) of a trait is known such as to estimate additive genetic variance ($\VarGenetic$) from phenotypic variance ($\VarPhenotype$) as $\VarGenetic = \Heritability \Multiply \VarPhenotype$.
Fortunately, if heritability is not known, the test for diversifying selection can still be performed, although it is underpowered.
Indeed, if the additive genetic variance is substituted by phenotypic variance, it is equivalent to assuming complete heritability ($\Heritability = 1$).
Because $\Heritability \leq 1$ by definition, we overestimate the within-species variation and thus underestimate $\NI$.
It is, however, possible to test for diversifying selection because testing for $\NI > 1$ while using phenotypic variance instead of additive genetic variance means that knowing the additive genetic variance would have only increased the evidence for diversifying selection.
Similarly, using the broad-sense heritability ($H^2$) instead of narrow-sense heritability ($\Heritability$) results in an underestimation of $\NI$ since $\Heritability \leq H^2$ and thus can be used to detect diversifying selection if $\Heritability$ is not available.
Additionally, empirical estimates of $\Heritability$ are surprisingly stable across species and fall within the range of 0.2-0.5 in a vast majority of phenotypic traits tested~\citep{hansen_heritability_2011, hansen_evolvability_2021}.
Thus, if available, such prior knowledge on $\Heritability$ can be leveraged instead of assuming complete heritability to increase the statistical power to detect diversifying selection.

In contrast to the test of diversifying selection, the test for stabilizing selection is invalid if $\NI$ is underestimated.
Several assumptions made by our test might not hold on empirical data and their consequences on the neutrality index and the test that can be performed are shown in Table~\ref{table:assumptions}.

\subsection*{Simulation}\label{subsec:simulations}

% vG: 13.8, vP: 69.0, vE: 55.2, nbr_loci: 5000, a: 1.0, mut_rate: 1.38e-05, pop_size: 50, h2: 0.2
% pS = 0.0027577478571428568
% Tree length (dS) = 2.340579
% Tree length (My) = 1259.4490200000002
% Root age (My) = 98.9489413888889
% Root age (dS) = 0.18388820079995308
% nbr_sites_var = 27.577478571428568
% u = 1.3788739285714283e-05
% nbr_generations = 13336.114128321291
We tested the performance of our neutrality index ($\NI$) to detect selection on a quantitative trait using simulations.
We performed simulations under different selective regimes (neutral, stabilizing, diversifying), different demographic histories (constant or fluctuating population size) and different evolution of the mutation rate (constant or fluctuating).
Simulations were individual-based and followed a Wright-Fisher model with mutation, selection and drift for a diploid population including speciation along a predefined ultrametric phylogenetic tree (Fig.~\ref{fig:simulator}A\&B).
Each individual phenotypic value was the sum of genotypic value and an environmental effect.
The environmental effect was normally distributed with variance $\VarEnv$.
We assumed that the genotypic value was encoded by $\NbrLoci=5,000$ loci, with each locus contributing an additive effect that was normally distributed with standard deviation $a=1$ (Fig.~\ref{fig:simulator}A and for the theoretical formulation see section~S\ref{subsec:genotype-phenotype-map} and fig.~\ref{fig:simulator-summary}).
We assumed a trait with a narrow-sense heritability of $\Heritability=0.2$ and computed the theoretical $\VarEnv$ accordingly (see section~S\ref{subsec:genotype-phenotype-map}).
Assuming a diploid panmictic population of size $\Ne=50$ at the root of the tree, and with non-overlapping generations, we simulated explicitly each generation along an ultrametric phylogenetic tree.
For each offspring, the number of mutations was drawn from a Poisson distribution with mean $2 \Multiply \MutationRatePheno \Multiply \NbrLoci $, with the mutation rate per locus per generation $\MutationRatePheno$.
From the empirical mammalian dataset (see next section), we computed an average nucleotide divergence from the root to leaves of $0.18$ and average genetic diversity of $0.00276$.
We scaled parameters in our simulations to fit plausible values for mammals.
We thus used a nucleotide mutation rate of $\MutationRateNuc=0.00276 / 4 \Ne = 1.38 \times 10^{-5}$ per site per generation  and a total of $0.18 / 1.38 \times 10^{-5} = 13,500$ generations from root to leaves, and the number of generations along each branch was proportional to the branch length.
We set $\MutationRatePheno=\MutationRateNuc$ without loss in generality since the genetic architecture ($\NbrLoci$ and $a$) is assumed constant in the simulator.

The changes in $\MutationRatePheno$ and $\Ne$ along the lineages were both modeled by a Brownian process on the log scale (log-$\MutationRatePheno$ and log-$\Ne$), leading to geometric Brownian motion on the linear scale ($\MutationRatePheno$ and $\Ne$).
These processes are parameterized as $\brownian \left(0, \sigma_{\MutationRatePheno}=0.0086\right)$ and $\brownian \left(0, \sigma_{\Ne}=0.0086\right)$, which, if counted across $13,500$ generations, leads to a standard deviation of $0.0086 \Multiply \sqrt {13,500} = 1.0$.
In other words, the deviation in log-$\Ne$ and log-$\MutationRatePheno$  between the extant species and the root is $1.0$.
An Ornstein-Uhlenbeck process was overlaid to the instant value of log-$\Ne$ provided by the geometric Brownian process to account for short-term changes between generations ($\text{OU} \left(0, \sigma_{\Ne}=0.1, \theta_{\Ne}=0.9\right)$).
The geometric Brownian motion accounted for long-term fluctuations (low rate of changes $\sigma_{\Ne}$ but unbounded), while the Ornstein-Uhlenbeck introduced short-term fluctuations (high rate of changes $\sigma_{\Ne}$ but bounded and mean-reverting).
The simulation started from an initial sequence at equilibrium at the root of the tree and, at each node, the process was split until it finally reached the leaves of the tree.
From a speciation process perspective, this was equivalent to an allopatric speciation over one generation.

At each generation, parents were randomly sampled with a weight proportional to their fitness ($W$).
Selection was modeled as a one-dimensional Fisher's geometric landscape, with the fitness of an individual being a monotonously decreasing function of the distance between the individual and the optimal phenotype~\citep{tenaillon_utility_2014,blanquart_epistasis_2016}.
More specifically, the fitness of an individual was given by $W = \e^{(\Trait - \lambda)^2/ \alpha}$, where $\Trait$ was the trait value of the individual, $\lambda=0.0$ was the optimal trait value, and $\alpha=0.02$ was the strength of selection.
Mutations were considered as a displacement of the phenotype in the multidimensional space.
Beneficial mutations moved the phenotype closer to the optimum, while deleterious mutations moved it further away.
Stabilizing selection was implemented by fixing the optimum phenotype to a single value ($\lambda=0.0$).
Diversifying selection was implemented by allowing the optimum phenotype to move along the phylogenetic tree as a geometric Brownian process~\citep{hansen_stabilizing_1997} ($\lambda \sim \brownian \left(0, \sigma_{\lambda}=1.0\right)$).
Neutral evolution was implemented by flattening the fitness landscape ($W=1$), which meant that each individual had the same probability of being sampled at each generation.

Nucleotide diversity ($\pi$) was measured as the heterozygosity of neutral markers that were simulated along the phylogenetic tree but not linked to the trait simulated.
Nucleotide divergence ($d$) was measured as the number of substitutions per site of neutral markers along the branches of the phylogenetic tree.
The additive genetic variance was measured as phenotypic variance multiplied by heritability.
Heritability was estimated from the slopes of the regression of offspring's phenotypic trait values on parental phenotypic trait values~\citep{lynch_genetics_1998} averaged over the last 10 simulated generations.
Heritability was thus not a given parameter of the simulations, but rather measured as it would be in empirical data.

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\textwidth, page=1] {figure2}
    \caption{
        Wright-Fisher simulations with mutation, selection and drift.
        Left panel: For a given individual, the trait phenotypic value is the sum of genotypic value and a environmental effect (standard deviation $\VarEnv$).
        The trait's genotypic value is encoded by $\NbrLoci$ independent loci (meaning no linkage), with each locus contributing additively to the genotypic value.
        Parents are selected for reproduction to the next generation according to their phenotypic value, with a probability proportional to their fitness.
        Mutations are drawn from a Poisson distribution, with each locus having a probability $\MutationRatePheno$ to mutate.
        Drift is modeled by the resampling of parents.
        Right panel: examples of a trait evolving along a phylogenetic tree, with the mean phenotype (black line) and the variance of the trait genotypic value (gray area).
    }
    \label{fig:simulator}
\end{figure*}

>>>>

\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{1.2}
\frenchspacing

@@

\section{Bayesian implementation}\label{sec:implementation}

Bayesian implementation is included within the \textit{BayesCode} software, available at \url{https://github.com/ThibaultLatrille/bayescode}.
